---
title: "Independent t-test"
authors:
- admin
date: 2022-04-14T21:13:14-05:00
output:
  html_document:
    highlight: textmate
categories: ["R"]
subtitle: Adopted from Handbook of Univariate and Multivariate Data Analysis with IBM SPSS by Robert Ho.
summary: The independent *t* test is used for testing the difference between means of two independent groups. It is particularly useful when the research question requires the comparison of variables (measured at least at the ordinal level) obtained from two independent samples.
tags: ["t-test", "parametric", "compare means"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", comment = "")

library(tidyverse)
library(haven)
```

```{r echo=FALSE}
# Loading data
memory_ability <- 
read_sav("https://raw.githubusercontent.com/chris-allones/Research-methods/master/t-test/data/EX4.sav") %>% 
     mutate(gender = as.character(gender),
            gender = factor(gender, levels = 1:2, labels = c("Male", "Female")))

mean_words_male <- 
memory_ability %>% 
     filter(gender == "Male") %>% 
     summarise(mean_word = mean(words)) %>% 
     pull()

mean_words_female <- 
memory_ability %>% 
     filter(gender == "Female") %>% 
     summarise(mean_word = mean(words)) %>% 
     pull()

```

```{r include=FALSE}
memory_ability %>% 
     ggplot(aes(x = words, fill = gender,color = gender)) +
     geom_density(alpha = 0.3, size = 0.8) +
     geom_vline(xintercept = mean_words_male, lty = "dashed", color = "red") +
     geom_vline(xintercept = mean_words_female, lty = "dashed", color = "cyan4") +
     scale_x_continuous(limits = c(10, 35)) +
     theme(legend.position = "top") +
     annotate(geom = "text", label = "Mean no of words \nrecalled: 17.7",
              x = 18, y = 0.145, hjust = 0, size = 3.5, color = "red") +
     annotate(geom = "text", label = "Mean no of words \nrecalled: 22.1",
              x = 22.3, y = 0.13, hjust = 0, size = 3.5, color = "cyan4")

```

<br>

# Aim

The independent *t* test is used for testing the difference between means of two independent groups. It is particularly useful when the research question requires the comparison of variables (measured at least at the ordinal level) obtained from two independent samples.

<br>

# Assumptions

+ Independence - the two groups are independent of one another.

+ Normality - the dependent variable is normally distributed.

+ Homogeneity of variance - the distribution of the dependent variable for one of the groups being compared has the same variance as the distribution for the other group being compared.

<br>

# Data

```{r}
skimr::skim_tee(memory_ability)
```

<br>

# Testing assumptions

## Independence

During data collection, ensure that the observation in one group are independent of the observations of the other group.

<br>

## Normality test

For a start, a qqplot or the "quantile-quantile" plot can be used to quickly assess whether or not the data or a variable follows a normal distribution. If normally distributed, the data points will lie on the straight diagonal line. For the qqplot below, most of the points lie on the diagonal line but with some deviations along each of the tails. 

```{r}
qqnorm(memory_ability$words)
qqline(memory_ability$words)
```

Whereas the Shapiro-Wilk statistic tests for normality. If the significance level are greater than 0.05, then normality is assumed. The Shapiro-Wilk test is recommended when sample size is small (~ <50). Result of the Shapiro-Wilk test below, the computed significance level is 0.201, implying normality can be assumed.

```{r}
shapiro.test(memory_ability$words)
```

<br>

## Homogeneity of variance

The Levene's test can be used to check the assumption on the homogeneity of variance. The Levene's tes for equality of variance tests the hypothesis that the two population variances are equal. 

The results below shows that the Levene statistic is *F=0.087* and the corresponding level of significance is large (i.e., *p>0.05*). Hence, the assumption of homogeneity of variance has not been violated, and the equal variances assumed implying *t* test statistic can be used for evaluating the null hypothesis of equality of means.

```{r}
car::leveneTest(words ~ gender, data = memory_ability, center = "mean")
```

<br>

# Results and interpretation

The result from the *t* test indicate that there is a significant differenc between the male and female samples in the number of words correctly recalled, *t(df=18)=-3.02, p<0.01*. The mean values indicate that females correctly recalled significantly more words (M=22.10) than males(M=17.70).

The confidence interval information shows that the null hypothesis value (i.e., zero) does not fall within the interval. Therefore the null hypothesis of equality of means can be rejected.

```{r}
ind_ttest <- 
t.test(words ~ gender, data = memory_ability, alternative = "two.sided")

ind_ttest

```

We can also compute for the eta-square ($\eta^2$) which can be interpreted in exactly the same as $R^2$ in correlation and regression. That is, approximately 33.63% of the variability in the number of words correctly recalled was explained by the gender manipulation. Eta-square can be calculated by:

$$\eta^2 = \frac{t^2}{t^2 + df}$$

```{r}
ind_ttest$statistic^2 / (ind_ttest$statistic^2 + ind_ttest$parameter)
```



